Characterizing subarctic European vegetation for land surface modeling of ozone pollution and climate risk

### 2021-09-27 VegOzone.tex
Add  changes from Ane to Section 4 and 5

The estimate of leaf boundary layer resistance is a function of wind speed which will be influenced by the tree height and leaf dimensions. Leaves of downy birch in northern Fennoscandia are probably smaller than the $5\,\unit{cm}$ from the MM parameterization of silver birch leaves. A sample of downy birch leaves collected at Svanhovd had an average length of $(3.0\pm 0.5)\,\unit{cm}$. In addition, \citet[][p.~52]{NINA2004} indicate an average tree height of $13.5\,\unit{m}$ in the area. Both factors will affect leaf boundary layer resistance, the smaller leaves would tend to reduce leaf boundary layer resistances whilst the shorter trees would reduce canopy roughness and increase canopy boundary layer resistances and hence the uptake of \chem{O_3}. In addition, one Scots pine forest at Svanhovd has been studied as a part of the ICP Forest mapping. Heights were measured in 2004, when the stand was 90 years old. Mean tree height was $10.1\,\unit{m}$ and maximum height was $16.2\,\unit{m}$ \textbf{TODO:(Volkmar Timmerman, priv. comm.)}. Canopy height and leaf size are adjusted to local values in the models with bespoke growing season (GS bespoke) in the following.
%%%
\subsection{$\mathrm{POD_1}$ results and implications on ozone impact}
\label{subsec:do3se_results}

We have modeled $\mathrm{POD_y}$ with a flux threshold $y=1\,\unit{nmol\,m^{-2}\,s^{-1}}$ per projected leaf area (PLA) for the three natural/semi-natural vegetation types, deciduous trees, coniferous trees, and perennial grassland using the generic MM parameterization and our bespoke temperaure (\emph{cold}, \emph{subarctic}) and light (PPFD0.8, PPFD1.2) acclimation types, and localized GS. The results are shown in Fig.~\ref{fig:pody_rel}. For comparison, critical levels for a $4\,\unit{\%}$ (deciduous), $2\,\unit{\%}$ (coniferous) \citep[refer to silver birch and Norway spruce in][]{ICP:MappingManual2017}, and $10\,\unit{\%}$ (perennial grassland, above ground) \citep{ESPR:Hayes2021} reduction in biomass are shown as horizontal dashed lines. 2018 is color-coded in pink and 2019 in purple. Simulations with MM GS, canopy, and leaf parameters are indicated with squared markers, bespoke GS with circles. The annotation $\pm$ refers to the bespoke light response function ($+\rightarrow\mathrm{PPFD1.2}$, $-\rightarrow\mathrm{PPFD0.8}$), while bespoke temperature response functions are indicated by categories on the x-axis. Closed/open symbols represent SWP sensitivity switched off/on in the model.

The results depicted in Fig.~\ref{fig:pody_rel} allow for a comprehensive assessment of a large variety of systematic uncertainties in the exposure-based risk metric for ozone induced damage in subarctic vegetation. In summary, we identify the following main effects of the bespoke parameterizations of the $\mathrm{DO_3SE}$ model with respect to the MM default parameterization on ozone uptake:
\begin{itemize}
  \itemsep0pt
\item Increasing cold tolerance/heat intolerance $\Rightarrow$ increased uptake in both probed years,
\item usage of proper growing season $\Rightarrow$ reduced uptake with respect to the increase reported above,
\item an earlier assumed start of GS $\Rightarrow$ increased uptake,
\item drought conditions (SWP) $\Rightarrow$ reduced uptake,
\item lower/higher extent of stomatal opening at low light intensities $\Rightarrow$ increased/reduced uptake,
\item symmetric variation of the extent of stomatal opening at low light intensities $\Rightarrow$ asymmetric response in POD,
\item temperature acclimation $\Rightarrow$ amplifying effect on light threshold and drought effects.
\end{itemize}
The exact order of magnitude of these effects varies for individual species or PFTs and between years, but the predicted ozone uptake for the bespoke parameterization is always larger than for the MM default parameterizations and of the same order of magnitude as the variability between the years studied here. Although our temperature acclimations are hypothetical, they suggest considerable underestimations of ozone risk for subarctic species when compared to generic MM parameterizations. In particular, an earlier GS due to climate change, that is overlapping with the ozone spring peak, has the potential to increase the risk of ozone damage for vegetation with an acclimation to cold climates. Drought effects (SWP) do only matter if a bespoke GS is not taken into account. We will look at the specific species or PFTs in more detail in the following.

For deciduous trees (Fig.~\ref{fig:pody_rel}a) all simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Deciduous trees display the largest variance of ozone uptake among all species or PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light threshold is especially pronounced, e.g. $\Delta\mathrm{POD_1} = ^{+2.5}_{-3.6}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of the subarctic type without bespoke GS in 2019. Without bespoke GS, 2018 displays a higher ozone damage risk than 2019. However, taking the later start and overall shorter GS in 2018 into account, this is reversed for the \emph{subarctic} acclimation type. This can be partly explained by very high light response factors (over $f_\mathrm{light}>0.8$) in the beginning of the GS in these simulations. Whether this is a bug or a feature of the $\mathrm{DO_3SE}$ model is not clear. SWP induced reduction in ozone uptake is only effective in 2018 and without bespoke GS. The temperature acclimation dependent bias in ozone uptake deduced for bespoke GS ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum systematic deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of the extent of stomatal opening at low light intensities amount to $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

For coniferous trees (Fig.~\ref{fig:pody_rel}b), most simulations slightly exceed the CL ($0-10\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). Coniferous trees' acclimation to prevailing temperatures was already good from the beginning and therefore the least biased of the generic MM parameterization. Coniferous trees display a medium sensitivity to the variation in the light threshold, e.g. $\Delta\mathrm{POD_1} = ^{+1.7}_{-1.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in case of subarctic type without bespoke GS in 2018. Taking the bespoke GS into account, we find a higher ozone uptake in 2019 for all bespoke temperature response functions. The overall bias deduced from bespoke GS for subarctic and cold types with respect to MM ranges between $(0.5-1.0)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.0-3.3)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum systematic deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of the extent of stomatal opening at low light intensities amount to $^{+1.0}_{-0.8}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+1.3}_{-1.0}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

For perennial grassland (Fig.~\ref{fig:pody_rel}c), all simulations with bespoke GS stay below the CL. Perennial grassland shows the smallest ozone uptake, but a similarly large bias as deciduous trees. We also find a reversal for predicted ozone damage risk in $\mathrm{POD_1}$ between 2018 and 2019 for the subarctic type. Perennial grassland shows the lowest sensitivity to the light threshold and SWP is only relevant without bespoke GS. The overall bias deduced from bespoke GS for subarctic and cold types with respect to MM ranges between $(2.2-3.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(4.4-6.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum systematic deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of the extent of stomatal opening at low light intensities amount to $^{+0.3}_{-0.2}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+0.3}_{-0.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

\begin{figure}[t]
  \includegraphics[width=8.3cm]{fig09}
  \caption{$\mathrm{DO_3SE}$ modeling $\mathrm{POD_1}$ results for (a) deciduous trees, (b) coniferous trees, and (c) perennial grasslands at Svanhovd in 2018 (pink) and 2019 (purple). The results varied depending on parameterizations of temperature response functions (MM, Cold, and Subarctic, details in Table~\ref{tab:sensitivity_tests_temp}), growing season (MM (squares) or bespoke (circles), details in Table~\ref{tab:sensitivity_tests_gs}), light response functions (first, second and third point attached by a line are depicting the MM, PPFD0.8 and PPFD1.2, details in Table~\ref{tab:sensitivity_tests_light}), and the effect of taking SWP into account (filled symbols without and open symbols with effects of drought kept in the model). Critical levels for ozone damage \citet{ICP:MappingManual2017,ESPR:Hayes2021} are given as dashed horizontal lines. Note that the horizontal axis is only categorical.}
  \label{fig:pody_rel}
\end{figure}

We estimated the biomass reductions in accordance to the CLs in \citet{ICP:MappingManual2017, ESPR:Hayes2021} for the most extreme temperature acclimation (\emph{subarctic}) for each bespoke PFT. The estimates are listed in Table~\ref{tab:biomass_reduction}. The reported range in uncertainty corresponds to the maximum deviation defined by the $\pm 20\,\unit{\%}$ variation of the extent of stomatal opening at low light intensities with SWP=on. For deciduous trees and perennial grassland, substantial reductions in total biomass in comparison with the default generic MM parameterizations are predicted for both years. Coniferous trees are the least affected and the magnitude of total biomass reduction is independent of the choice of the parameterization within given uncertainties. This may reflect the more comprehensive research on this species or PFT feeding into the generic MM parameterization. The systematic deviation for deciduous trees is one order of magnitude larger than for coniferous trees and perennial grassland.

From the difference between 2018 and 2019, we infer a substantial interannual variability. The interannual variability is estimated to be larger than our reported systematic deviations and can be of the same order of magnitude as the effect of the temperature acclimation. This means that the choice of temperature acclimation has the potential to either revoke or amplify the risk of ozone damage for vegetation. However, the risk of ozone damage in deciduous trees and perennial grassland in comparison with the generic MM parameterizations is increased, regardless.

\begin{table}[t]
  \caption{Estimated total biomass reduction in \unit{\%} for temperature acclimation \emph{subarctic} with bespoke GS and SWP=off. The uncertainty ranges reported here correspond to maximum a divergence deducted from varying the extent of stomatal opening at low light intensities and SWP=on. For comparison, the corresponding biomass reduction for the generic default MM averaged over both years is shown. The reported standard deviation is computed from all sensitivity simulations ($\pm 20\,\unit{\%}$ in extent of stomatal opening at low light, SWP on/off, and bespoke GS).}
  \label{tab:biomass_reduction}
\begin{tabular}{lllll}
\tophline
Year & \multicolumn{4}{c}{PFT}\\
& Deciduous tree& Coniferous tree & \multicolumn{2}{c}{Perennial grassland}\\
\middlehline
2018 & $15.5^{+(1.9...2.1)}_{-(0.8...1.5)}$ & $2.5^{+0.2}_{-0.2}$ & $9.7^{+0.2}_{-0.2}$ & $^*12.4^{+0.2}_{-0.2}$\\
2019 & $17.4^{+2.5}_{-1.8}$ & $3.0^{+0.2}_{-0.3}$ & $10.5^{+(0.1...0.2)}_{-(0.0...0.1)}$ & $^*13.7^{+(0.1...0.3)}_{-(0.1...0.3)}$\\
\middlehline
$\left<\mathrm{MM}\right>$ & $11.2\pm 1.1$ & $2.31\pm 0.04$ & $7.5\pm 0.9$ & $^*9.6\pm 1.4$\\
\bottomhline
\end{tabular}
\belowtable{$^*$ above ground biomass} % Table Footnotes
\end{table}

@@

We have studied the effect of acclimations of biomes to a subarctic climate on phyto-toxic ozone uptake. The comparison between 2018 and 2019 conditions allowed us to consider these acclimations and their influence on vegetation sensitivity to \chem{O_3} in light of future changes of key environmental variables as may occur under climate change (e.g. increase of frequency and extent of heatwaves). We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was particularly a result of the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden (in particular in mid--end July).

We found that the length and timing of the growing season are crucial for risk assessment based on accumulated ozone flux metrics ($\mathrm{POD_y}$). Based on a MODIS photosynthesis product, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The found start of growing season in 2018 ($122\,\unit{doy}$) and 2019 ($106\,\unit{doy}$) are within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. Based on gridded observational temperature and snow depth, we found a half a month shorter growing season for deciduous trees and perennial grassland in 2018 compared to 2019, respectively. The timing and emerging trend in the start of the growing season may differ substantially between local observation and common thermal or latitudinal estimations. Species-specific methods show a later start for European silver birch and an earlier for Norway spruce, compared to the thermal growing season \citep{IVL:Karlsson2018}. With respect to ongoing climate change, a clear positive trend in length ($5.2\,\unit{d\,decade^{-1}}$) almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) could be deducted from the thermal growing season at Svanhovd.

Ozone peak concentrations were observed more frequently in 2018 than in 2019. In 2018, we found significantly enhanced \chem{[O_3]} particularly in April and May, but not during the active wildfires in July. This suggests that weather conditions and high ozone concentration at the beginning of the growing season are more important in risk assessment using flux-based metrics than episodes of peak concentrations over shorter periods. Because these peak \chem{[O_3]} may coincident with environmental conditions that limit ozone uptake and hence do not lead to damage. Consequently, acute cellular damage caused by very high fluxes over a short period is not accounted for properly in $\mathrm{POD_y}$. As pointed out by \citet{AE:Musselman2006}, a higher temporal resolution, e.g. susceptibility to ozone depending on leaf age, in the associated damage functions might give a different picture. For herbal and grassland species (similar for crops) a strong dependency on the phenological state, e.g. high susceptibility to ozone in the reproductive state, was found \citep{EP:Bassin2004}.

>>>>

The estimate of leaf boundary layer resistance is a function of wind speed which will be influenced by the tree height and leaf dimensions. Leaves of downy birch in northern Fennoscandia are probably smaller than the $5\,\unit{cm}$ from the MM parameterization of silver birch leaves. A sample of downy birch leaves collected at Svanhovd had an average length of $(3.0\pm 0.5)\,\unit{cm}$. In addition, \citet[][p.~52]{NINA2004} indicate an average tree height of $13.5\,\unit{m}$ in the area. Both factors will affect leaf boundary layer resistance, the smaller leaves would tend to reduce leaf boundary layer resistances whilst the shorter trees would reduce canopy roughness and increase canopy boundary layer resistances and hence the uptake of \chem{O_3}. In addition, one Scots pine forest at Svanhovd has been studied as a part of the ICP Forest mapping. Heights were measured in 2004, when the stand was 90 years old. Mean tree height was $10.1\,\unit{m}$ and maximum height was $16.2\,\unit{m}$ \textbf{TODO:(Volkmar Timmerman, priv. comm.)}. Canopy height and leaf size are adjusted to local values in the models with bespoke growing season (GS bespoke) in the following.

\subsection{$\mathrm{POD_1}$ results and implications on ozone impact}
\label{subsec:do3se_results}

We have modeled $\mathrm{POD_y}$ with a flux threshold $y=1\,\unit{nmol\,m^{-2}\,s^{-1}}$ per projected leaf area (PLA) for the three natural/semi-natural vegetation types, deciduous trees, coniferous trees, and perennial grassland using the generic MM parameterization and our bespoke temperature (\emph{cold}, \emph{subarctic}) and light (PPFD0.8, PPFD1.2) response functions. We also locally adapted leaf size for deciduous trees and the tree height together with the bespoke GS. The results displayed in Fig.~\ref{fig:pody_rel} allow for an assessment of systematic uncertainties arising from the use of generic parameterizations established from less climatically extreme conditions.
For comparison, critical levels for reduction in biomass are shown as horizontal dashed lines \citep[deciduous forest $4\,\unit{\%}$, coniferous forest $2\,\unit{\%}$, and grasslands $10\,\unit{\%}$ biomass reduction,][]{ICP:MappingManual2017,ESPR:Hayes2021}. We identify that an increase in cold tolerance leads to higher ozone uptake in both probed years, while the usage of locally adapted GS causes a reduction. The earlier and longer GS in 2019 leads to increased ozone uptake compared to 2018 despite more episodes of elevated \chem{[O_3]} in 2018. Due to the shape of the light response function, a variation of the extent of stomatal opening at low light intensities leads to an asymmetric response in $\mathrm{POD_1}$ and can cause a higher sensitivity to drought conditions. Similarly, the temperature acclimation to cold or subarctic conditions can lead to an amplification of effects by droughts.

The magnitude of these effects varies between PFTs as well as years, but the predicted ozone uptake for the bespoke parameterization is always larger than for the MM default parameterizations and of the same order of magnitude as the variability between the years studied here. Although our temperature acclimations are hypothetical, they suggest considerable underestimations of ozone risk for subarctic species when relying on generic MM parameterizations. We will look at the specific PFTs in more detail in the following.

For deciduous trees (Fig.~\ref{fig:pody_rel}a) all $\mathrm{POD_1}$ simulations exceed the CL by $5-25\,\unit{mmol\,\chem{O_3}\,m^{-2}}$. Deciduous trees display the largest variation of ozone uptake among all PFTs ($10-31\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). In particular, the sensitivity to the light parameterization is pronounced. With MM GS, $\mathrm{POD_1}$ values were higher in 2018 than in 2019, throughout our analyses. However, taking the later start and overall shorter GS in 2018 into account (bespoke GS), this is reversed for the \emph{subarctic} acclimation type. This can be explained by favorable growing conditions in 2019. Low soil water potential reduced the $\mathrm{POD_1}$ values in 2018 in some cases when using the MM growing season, but did not or only slightly when using bespoke growing season. The temperature acclimation-dependent bias in ozone uptake deduced for bespoke GS ranges between $(1.4-3.6)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.7-7.7)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum systematic deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of the extent of stomatal opening at low light intensities amount to $^{+1.6}_{-2.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2018 and $^{+1.9}_{-2.7}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2019.

For coniferous trees (Fig.~\ref{fig:pody_rel}b), most simulations slightly exceed the CL ($0-10\,\unit{mmol\,\chem{O_3}\,m^{-2}}$). Coniferous trees' parameterization was already well acclimated to prevailing temperatures from the beginning and therefore the least biased of the generic MM parameterizations. Taking the bespoke GS and canopy height into account, we find a higher ozone uptake in 2019 than in 2018 for both bespoke temperature response functions. The overall bias deduced from bespoke GS for subarctic and cold types with respect to MM range between $(0.5-1.0)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(2.0-3.3)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The maximum systematic deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of the extent of stomatal opening at low light intensities is medium and amounts to $^{+1.0}_{-0.8}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2018 and $^{+1.3}_{-1.0}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ for 2019.

For perennial grassland (Fig.~\ref{fig:pody_rel}c), all simulations with bespoke GS stay below the CL. Perennial grassland shows the smallest ozone uptake, but a similarly large response to the temperature parameterizations as deciduous trees. Again, we find a reversal for predicted ozone damage risk in $\mathrm{POD_1}$ between 2018 and 2019 for the subarctic type. Perennial grassland shows the lowest sensitivity to the light threshold and SWP is only relevant without bespoke GS. The overall bias deduced from bespoke GS for subarctic and cold types with respect to MM ranges between $(2.2-3.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2018 and $(4.4-6.4)\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ in 2019. The lower maximum temperature tolerance in the subarctic compared to the cold temperature parameterization probably caused the lower $\mathrm{POD_1}$ values calculated for 2018 in subarctic compared to cold climate, whereas this was not seen for 2019. The maximum systematic deviations deduced from the $-20\,\unit{\%}$ (superscript) and $+20\,\unit{\%}$ (subscript) variation of the extent of stomatal opening at low light intensities amount to $^{+0.3}_{-0.2}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$ and $^{+0.3}_{-0.3}\,\unit{mmol\,\chem{O_3}\,m^{-2}}$, respectively.

\begin{figure}[t]
  \includegraphics[width=8.3cm]{fig09}
  \caption{$\mathrm{DO_3SE}$ modeling $\mathrm{POD_1}$ results for (a) deciduous trees, (b) coniferous trees, and (c) perennial grasslands at Svanhovd in 2018 (pink) and 2019 (purple). The results varied depending on parameterizations of temperature response functions (MM, Cold, and Subarctic, details in Table~\ref{tab:sensitivity_tests_temp}), growing season (GS MM (squares) or GS bespoke (circles), details in Table~\ref{tab:sensitivity_tests_gs}), light response functions (first, second and third point attached by a line are depicting the MM, PPFD0.8 and PPFD1.2, details in Table~\ref{tab:sensitivity_tests_light}), and the effect of taking SWP into account (filled symbols without and open symbols with effects of drought kept in the model). Critical levels for ozone damage \citep{ICP:MappingManual2017,ESPR:Hayes2021} are given as dashed horizontal lines. Note that the horizontal axis is only categorical.}
  \label{fig:pody_rel}
\end{figure}

We estimated the biomass reductions in accordance to the CLs in \citet{ICP:MappingManual2017, ESPR:Hayes2021} for the most extreme temperature acclimation (\emph{subarctic}) for each bespoke PFT. The estimates are listed in Table~\ref{tab:biomass_reduction}. The reported range in uncertainty corresponds to the maximum deviation defined by the $\pm 20\,\unit{\%}$ variation of the extent of stomatal opening at low light intensities with SWP=on. For deciduous trees and perennial grassland, total biomass was reduced in comparison with the default generic MM parameterizations for both years. \textbf{TODO: Discussion about CL for grassland!} Coniferous trees are the least affected and the magnitude of total biomass reduction is independent of the choice of the parameterization within given uncertainties. This may reflect that the temperature parameterization in the MM is almost never a limiting factor for stomatal opening. The systematic deviation for deciduous trees is one order of magnitude larger than for coniferous trees and perennial grassland.

\textbf{TODO: Check if the legnth effect central Europe vs Svanvik is propperly taken care of.}
From the difference between 2018 and 2019, we infer a substantial interannual variability. The interannual variability is estimated to be larger than our reported systematic deviations and can be of the same order of magnitude as the effect of the temperature acclimation. This means that the choice of temperature acclimation has the potential to either revoke or amplify the risk of ozone damage for vegetation. However, the risk of ozone damage in deciduous trees and perennial grassland in comparison with the generic MM parameterizations is increased, regardless.

\begin{table}[t]
  \caption{Estimated total biomass reduction in \unit{\%} for temperature acclimation \emph{subarctic} with bespoke GS and SWP=off. The uncertainty ranges reported here correspond to maximum a divergence deducted from varying $f_\mathrm{light}$ and SWP=on. For comparison, the corresponding biomass reduction for the generic default MM averaged over both years is shown. The reported standard deviation is computed from all sensitivity simulations ($\pm 20\,\unit{\%}$ in extent of stomatal opening at low light, SWP on/off, and bespoke GS).}
  \label{tab:biomass_reduction}
\begin{tabular}{lllll}
\tophline
Year & \multicolumn{4}{c}{PFT}\\
& Deciduous tree& Coniferous tree & \multicolumn{2}{c}{Perennial grassland}\\
& & & total & above ground\\
\middlehline
2018 & $15.5^{+(1.9...2.1)}_{-(0.8...1.5)}$ & $2.5^{+0.2}_{-0.2}$ & $9.7^{+0.2}_{-0.2}$ & $12.4^{+0.2}_{-0.2}$\\
2019 & $17.4^{+2.5}_{-1.8}$ & $3.0^{+0.2}_{-0.3}$ & $10.5^{+(0.1...0.2)}_{-(0.0...0.1)}$ & $13.7^{+(0.1...0.3)}_{-(0.1...0.3)}$\\
\middlehline
$\left<\mathrm{MM}\right>$ & $11.2\pm 1.1$ & $2.31\pm 0.04$ & $7.5\pm 0.9$ & $9.6\pm 1.4$\\
\bottomhline
\end{tabular}
%\belowtable{$^*$ above ground biomass} % Table Footnotes
\end{table}

@@

\textbf{TODO: Decide on terms acclimation vs parameterization etc} We have studied the effect of model acclimations of some common vegetation types to a subarctic climate on phytotoxic ozone uptake. The comparison between 2018 and 2019 conditions allowed us to consider these acclimations and their influence on vegetation sensitivity to \chem{O_3} in light of future changes of key environmental variables as may occur under climate change (e.g. increase of frequency and extent of heatwaves). We found that conditions for ozone formation were more favorable in the 2018 growing season than in 2019, with 2018 being significantly warmer and less cloudy in spring and early summer. Accordingly, peak ozone concentrations occurred more frequently and at higher levels in 2018. This was particularly a result of the extended heatwave in spring and early summer and associated, extensive wildfires in Central Sweden (in particular in mid--end July). \textbf{TODO: Check explanation regarding significance.}

We found that the length and timing of the growing season are crucial for risk assessment based on accumulated ozone flux metrics ($\mathrm{POD_y}$). Based on a MODIS photosynthesis product, we estimated the growing season for coniferous trees in 2018 to be at least one month shorter than in 2019. The found start of growing season in 2018 ($122\,\unit{doy}$) and 2019 ($106\,\unit{doy}$) are within given ranges by empirical observations \citep{TB:Kolari2007,IVL:Karlsson2018}. Based on gridded observational temperature and snow depth, we found a half a month shorter growing season for deciduous trees and perennial grassland in 2018 compared to 2019, respectively. With respect to ongoing climate change, a clear positive trend in length ($5.2\,\unit{d\,decade^{-1}}$) almost equally distributed between earlier start ($2.9\,\unit{days\,decade^{-1}}$) and later end ($2.3\,\unit{d\,decade^{-1}}$) could be deducted from the thermal growing season at Svanhovd (Supplement Fig.~\ref{fig:greening_season_change_Svanvik}). \textbf{TODO: Move S6 into appendix.}

Ozone peak concentrations were observed more frequently in 2018 than in 2019. Nevertheless, \chem{POD_1} values were very similar for both years. This suggests that the overall length of the growing season is more important in risk assessment using flux-based metrics than episodes of peak concentrations. These peak \chem{[O_3]} often coincide with environmental conditions that limit ozone uptake the model. However, acute cellular damage caused by very high episodic ozone fluxes are not accounted for in $\mathrm{POD_y}$. That might explain why ozone damage was observed at Svanhovd to a larger extent in 2018 than in 2019. As pointed out by \citet{AE:Musselman2006}, a susceptibility to ozone depending on leaf age in the associated damage functions might give a different picture. Forbs and perennial grassland have been found to be more susceptible in the reproductive state \citep{EP:Bassin2004}.
